---
title: "Exploring Age and Sex Dynamics in Juvenile Violence Trends"
author: "Nathan Rethwisch"
date: "2024-05-07"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Use echo = FALSE for Portfolio assignments
knitr::opts_chunk$set(echo = FALSE)
```

```{r metadata, echo=FALSE}
# Author:  Nathan Rethwisch
# Date:    2024-05-07
# Purpose: Analyze how sex and age affect proportion of violent crimes among those in juvenile detention
#-------------------------------------------------------------------------------
```

```{r packages, warning = FALSE}
suppressPackageStartupMessages(library("tidyverse")); theme_set(theme_bw())
suppressPackageStartupMessages(library("knitr"))
suppressPackageStartupMessages(library("flextable"))

```

The purpose of this document is to analyze how sex and age affect the
proportion of juvenile offenders that were detained in a juvenile
facility for violent crimes as opposed to nonviolent crimes. The
document highlights where the data comes from, the procedure and results
used to model the effect of sex and age on violence, and an overall
discussion of the conclusions drawn from these results.

## Introduction

```{r dataCleaning, echo = FALSE, warning = FALSE, message = FALSE}
#Reading the data and making violent crimes a count
juvenile <- read_csv("IowaJuvenileData.csv")
juvenile <- juvenile %>% mutate(violent = if_else(`Charge Type` == "VIOLENT", 1, 0))
```

Understanding trends among juvenile violence is important for assessing
risk factors, creating new policy, and advocating for reform within the
juvenile legal system. Therefore, in this study, we will take a further
look at how age and sex are related to the proportion of juveniles in
detention facilities that have committed violent crimes.

This data comes from the Iowa government and highlights all juvenile
delinquency placements from July 2003 through October 2020. The dataset
includes information on the child's age when admitted, type of charge
they were admitted on, the sex and race of the child, the facility they
were kept at, the month and year of their release, and the number of
days they were held. In total, there were `r nrow(juvenile)[1]` juvenile
placement records in this database.

More information on the dataset can be found at the [Iowa Data
website.](https://data.iowa.gov/Juvenile-Detention/Iowa-Juvenile-Detention-Secure-Placements/3kck-gq3y/about_data)

## Data

```{r violentNonviolent, include = FALSE}
#Getting seperate datasets for violent and nonviolent crimes
violent    <- juvenile%>%filter(`Charge Type` == "VIOLENT")
nonviolent <- juvenile%>%filter(`Charge Type` != "VIOLENT")
```

For this analysis, we want to focus on the proportion of crimes that
were considered violent among those placed in a juvenile detention
center. [According to Iowa
Legislature](https://www.legis.iowa.gov/docs/code/915.10.pdf), a violent
crime includes "a forcible felony, or any felony or aggravated
misdemeanor which involved the actual or threatened infliction of
physical or emotional injury on one or more persons." In this dataset,
`r nrow(violent)[1]` of the `r nrow(juvenile)` juveniles were detained
for violent crimes, meaning that
`r round(nrow(violent)[1] / nrow(juvenile) * 100, digits = 2)`% of
juveniles in detention centers are there for violent crimes. These
mainly included arson, assault, kidnapping, murder, burglary/theft, and
sex crimes. The table below gives a breakdown of violent crimes by what
type of crime was committed:

```{r subtype_counts, ft.align="left"}

# Define the subtypes to be combined into "OtherRobbery"
Burglary   <- c("BURGLARY", "ROBBERY", "THEFT")

# Define the subtypes to be combined into "Other"
Other      <- c("OTHER CRIMINAL", "OTHER PUB ORD", "OTHER VIOLENT")

Murder     <- c("MURDER/MANSL.")
Kidnapping <- c("KIDNAP")

# Replace the specified subtypes with "OtherRobbery" and "Other"
violent <- violent %>%
  mutate(`Charge Subtype` = case_when(
    `Charge Subtype` %in% Burglary ~ "Burglary/Robbery",
    `Charge Subtype` %in% Other ~ "Other",
    `Charge Subtype` %in% Murder ~ "Murder/Manslaughter",
    `Charge Subtype` %in% Kidnapping ~ "Kidnapping",
    TRUE ~ `Charge Subtype`
  ))

#Updating so subtypes aren't in all-caps
violent$`Charge Subtype` <- str_to_title(violent$`Charge Subtype`)

# Summarize counts for each subtype
subtype_counts <- violent %>%
  group_by (`Charge Subtype`) %>%
  summarise(Count = n())%>%
  mutate   (`Total Violent Crimes` = sum(Count),
            Proportion = round(Count / `Total Violent Crimes` * 100, 1),
            `Count (Proportion)`   = paste0(Count, " (", Proportion, "%)")) %>%
  arrange  (desc(`Count`))%>%
  select   (`Charge Subtype`, 
            `Count (Proportion)`)

#Creating a footer for the table
footer <- data.frame(col_keys = c("Charge Subtype", "Count (Proportion)"), 
                     unit     = c("Total", nrow(violent)[1]))


# Outputting the flextable with the footer and other formatting
ft_1 <- flextable(subtype_counts)
ft_2 <- set_footer_df(ft_1, mapping = footer, key = "col_keys")
ft_2 <- theme_booktabs(ft_2)
ft_2 <- fix_border_issues(ft_2)
ft_2 <- width(ft_2, width = 1.5)
ft_2

```

```{r sexTable, message = FALSE}
#Total offenders by sex
allCrimesSex     <- juvenile %>% group_by (Sex)%>%
                                 summarise(`Total Offenders`   = n())

#Violent offenders by sex
violentCrimesSex <- juvenile %>% group_by (Sex)%>%
                                 filter   (`Charge Type`       == "VIOLENT")%>%
                                 summarise(`Violent Offenders` = n())

#Creating a table with proportion information
sexTable <- allCrimesSex%>%left_join(violentCrimesSex)
sexTable <- sexTable%>%
      mutate(
             `Proportion Who are Violent Offenders` = round(`Violent Offenders` / `Total Offenders`, digits = 2),
              Proportion                            = round(`Total Offenders` / sum(`Total Offenders`) * 100, digits = 2),
             `Total Offenders`                      = paste0(`Total Offenders`, " (", Proportion, "%)")
            )%>%
      select(Sex, `Total Offenders`, `Violent Offenders`, `Proportion Who are Violent Offenders`)
```

Furthermore, there were more males that were in juvenile detention
centers in this dataset, with
`r count(juvenile %>% filter(Sex == "MALE"))[[1]]` males and only
`r count(juvenile %>% filter(Sex == "FEMALE"))[[1]]` females. The
following table breaks down the number and proportion of violent crimes
by gender:

```{r outputSexTable}
knitr::kable(sexTable, format = "markdown", row.names = FALSE, align = 'c')
```

Although there are far more males in the dataset, with
`r round(table(juvenile$Sex)[2] / nrow(juvenile)[1] * 100, digits = 2)[[1]]`%
of detained juveniles in Iowa being male, females were actually more
likely to be detained on violent crimes, with `r sexTable[[1,4]] * 100`%
of female juveniles committing violent crimes and
`r sexTable[[2,4]] * 100`% of male juveniles committing violent crimes.

```{r ageProp}
#This wouldn't run inline so I had to run this in a chunk
ageProp <- round(sum(table(violent$`Admission Age`)[7:9]) / nrow(violent)[1] * 100, digits = 2)
```

When looking at the ages, the bulk of admissions into juvenile detention
centers seems to be between the ages of 15-17, with `r ageProp`% of
juvenile offenders being between those ages. The following table breaks
down the number and proportion of violent crimes by age:

```{r ageTable, message = FALSE}
#Total offenders by age
allCrimesAge     <- juvenile %>% group_by (`Admission Age`)%>%
                                 summarise(`Total Offenders`   = n())

#Violent offenders by age
violentCrimesAge <- juvenile %>% group_by (`Admission Age`)%>%
                                 filter   (`Charge Type`       == "VIOLENT")%>%
                                 summarise(`Violent Offenders` = n())

#Creating a table with proportions of offenses
AgeTable <- allCrimesAge%>%left_join(violentCrimesAge)
AgeTable <- AgeTable%>%
  mutate(
         `Proportion Who are Violent Offenders` = round(`Violent Offenders` / `Total Offenders`, digits = 2),
          Proportion                            = round(`Total Offenders` / sum(`Total Offenders`) * 100, digits = 2),
         `Total Offenders`                      = paste0(`Total Offenders`, " (", Proportion, "%)")
         )%>%
  select(`Admission Age`, 
         `Total Offenders`,
         `Violent Offenders`, 
         `Proportion Who are Violent Offenders`)
```

```{r outputAgeTable}
knitr::kable(AgeTable, format = "markdown", row.names = FALSE, align = 'c')
```

Note that 18 and 19 year olds can be selected for juvenile detention if
they are deemed "mentally immature," so this explains the few offenders
in these age ranges. Interestingly, there seems to be a linear
relationship between age and the proportion of juveniles who are violent
offenders, with 100% of 9-year olds being sentenced for violent crimes,
and decreasing proportions as the ages increase. This decrease
intuitively makes sense because younger children are likely to be given
lesser punishments as long as the crimes are nonviolent. The proportion
slightly increases for 18 and 19 year-olds, but there are few total
offenders due to most people this age being sentenced to adult detention
centers.

The figure below displays a scatterplot, with the proportion of violent
crimes on the y-axis, and the age when admitted to the juvenile
detention facility on the x-axis. Furthermore, the sex of the child is
denoted by both the color and symbol. There were no female offenders in
the dataset that were age 9 or 19, so this is why there are no points on
the graph for those ages.

```{r juvenile_grouped, message = FALSE}
#Creating a grouping of all age-sex combos
juvenile_grouped <- juvenile%>%group_by(Sex, `Admission Age`)%>%
  summarise(number                   = n(),
            violent                  = sum(violent == 1),
            `Proportion of Violence` = violent / number)

```

```{r exploratory}
#Scatterplot of Violent Offenes by Age/Sex
ggplot(juvenile_grouped,
       aes(
         x     = `Admission Age`,
         y     = `Proportion of Violence`,
         color = Sex,
         shape = Sex
       )) +
  geom_point(
    position   = position_jitterdodge(
      dodge.width  = 0.1,
      jitter.width = 0.1
    ),
      size         = 2
    ) +
  labs(
    x     = "Age When Admitted",
    y     = "Proportion of Violent Offenses",
    title = "Proportion of Violent Offenses Based on Age and Sex"
  )
```

This plot shows that, in general, the proportion of violent offenders
seems to decrease as age increases. However, it appears that sex has
little to do with proportion of violent offenders. There is a deviation
for 18-year olds, but this is likely due to a lack of data, as only
there were only `r juvenile_grouped$number[20]` male offenders and
`r juvenile_grouped$number[9]` female offenders.

## Results

```{r model-interaction}

#Creating the models with and without an interaction term
model <-  glm(violent ~ Sex + `Admission Age`, 
              data   = juvenile, 
              family = 'binomial')

modelInteraction <- glm(violent ~ Sex + `Admission Age` + Sex:`Admission Age`, 
                        data   = juvenile, 
                        family = 'binomial')

#Had to include this because it couldn't be computed inline
interactionPvalue <- round(summary(modelInteraction)$coefficients[[4,4]], digits = 2)
```

Because we are trying to model the probability of something occurring, a
logistic regression model is appropriate in this scenario. In this case,
the response variable is the probability of a child in juvenile
detention being there for a violent crime. The explanatory variables are
sex and age when admitted. An interaction term between admission age and
sex was considered, but was not included in the final model, because it
was not found to be significant from a Wald test (P-value of
`r round(summary(modelInteraction)$coefficients[[4,4]], digits = 2)`).

The binomial model looks like this:
$$\hat{p} = \frac{e^{B_0 + B_1X_1 + B_2X_2}}{1+e^{B_0 + B_1X_1 + B_2X_2}}$$where

-   $\hat{p}$ is the predicted probability of the juvenile
-   $B_0$ is the intercept
-   $B_1$ is the coefficient associated with Sex
-   $X_1$ is whether the observed juvenile is male (1) or female (0)
-   $B_2$ is the coefficient associated with Age
-   $X_2$ is the age of the observed juvenile

After fitting our data, we get a model that looks like this:
$$\hat{p} = \frac{e^{`r model$coefficients[[1]]` + `r model$coefficients[[2]]` * X_1 + `r model$coefficients[[3]]` * X_2}}{1 + e^{`r model$coefficients[[1]]` + `r model$coefficients[[2]]` * X_1 + `r model$coefficients[[3]]` * X_2}}$$.

```{r confints, message = FALSE}
Sex <- exp(confint(model))[2,]
Age <- exp(confint(model))[2,]
```

After performing a Wald test on the model, we find that both sex and
admission age are significant predictors for the proportion of violent
offenders, given the other predictors in the model. From this, we can
obtain a 95% confidence interval to see what impact sex and age may have
on the proportion of offenders in juvenile detention that have committed
a violent crime.

A 95% confidence interval for the sex being female is
(`r round(1 / Sex[2],3)`, `r round(1 / Sex[1],3)`). This means that we are
95% confident that the odds of a female being a violent offender given
they are in juvenile detention are between `r round(1 / Sex[2],3)` and
`r round(1 / Sex[1],3)` times the odds of a male being a violent offender.

A 95% confidence interval for the admission age is
(`r round(Age[1], 3)`, `r round(Age[2], 3)`). This means that we can be
95% confident that the odds of a one-year increase in age is associated
with between a $\frac{1}{`r round(Age[2], 3)`}$ = `r round(1 / Age[2], 3)`
and $\frac{1}{`r round(Age[1],3)`}$ = `r round(1 / Age[1], 3)` times
*decrease* in the odds of juvenile delinquent being a violent offender.

The following plot shows the model's prediction for the proportion of
children in juvenile detention that are there for a violent crime based
on levels of admission age and sex as a line, with an associated 95%
confidence interval. Because there are so many observations in this
data, we have chosen to have the plotted points represent average
proportions among the unique combinations of age and sex.


```{r predictions}
#Expit function for later confidence intervals
expit <- function(x) {
  1 / (1 + exp(-x))
}

#Creating a grid for predictions
nd    <- expand.grid(
                    Sex             = unique(juvenile$Sex),
                    `Admission Age` = unique(juvenile$`Admission Age`)
                    )%>%
             arrange(desc(Sex), `Admission Age`)
```

```{r graphing}
#Predictions based on the model
p <- bind_cols(
  nd,
  predict(model,
          newdata = nd,
          se.fit  = TRUE)%>%
      as.data.frame() %>%
      
      # Manually construct confidence intervals
      mutate(
        lwr = fit - qnorm(0.975) * se.fit,
        upr = fit + qnorm(0.975) * se.fit,
        
        # Expit to get to response scale
        Violent    = expit(fit),
        lwr        = expit(lwr),
        upr        = expit(upr)
      ) 
  )

#Ensures the coloring for Sex remains consistent among all plotting methods
p$Sex <- factor(p$Sex, levels = c("FEMALE", "MALE"))


g <- ggplot(juvenile_grouped, 
         aes(x     = `Admission Age`,
             y     = `Proportion of Violence`,
             color = Sex,
             shape = Sex
             )) +
  geom_point(
      position = position_jitter(
            width = 0.1
            )) 


#Adding line for predicted probabilities
gm <- g +
  geom_line(
    data     = p,
    mapping  = aes(
      y      = Violent)) +
  labs(
    x        = "Admission Age (Years)",
    y        = "Proportion of Juveniles that Are Detained for Violent Crimes",
    title    = "Predicted Proportion of Violent Crimes Based on Admission Age and Sex",
    subtitle = "Logistic Regression Model with 95% Confident Intervals"
  )


#Adding pointrange for confidnece interval
g1 <- gm + geom_ribbon(
  data    = p,
  mapping = aes(
    y     = Violent,
    ymin  = lwr,
    ymax  = upr,
    fill  = Sex
  ),
  color   = NA,
  alpha   = 0.2
)

g1
```

From this plot, we can tell that females are predicted to have a
slightly higher proportion of being detained for violent crimes and age
is related to a decrease in the proportion of being admitted for violent
crimes.

All of this information leads to the conclusion that males have a lower
proportion of children in juvenile detention for violent crimes than
females. Furthermore, as age increases, the proportion of children in
juvenile detention for violent crimes decrease. It is important to note
that a decreased proportion does not equate to a decrease in the overall
count of children in juvenile detention for violent crimes, as older
males have a higher number of violent crimes, but a lower proportion.

## Discussion

This analysis was to ascertain the effect of sex and age on the
proportion of juveniles that were sent to a detention facility for
violent crimes. The data used for this analysis was all juvenile
detention admissions in the state of Iowa from July 2003 through October
2020, taken from [Iowa's government
website](https://data.iowa.gov/Juvenile-Detention/Iowa-Juvenile-Detention-Secure-Placements/3kck-gq3y/about_data).
After fitting a logistic regression model, we found that females have a
higher proportion of violent admissions than males and that the
proportion of admitted children who committed violent crimes decreases
with age.

Some of the assumptions of using a logistic regression model were
challenged in this analysis. One assumption is that observations have a
binomial distribution, each with its own mean. The graph in the
*Results* sections proves this assumption. It shows that the expected
proportions from a binomial distribution are close to the actual
proportions of violent offenders. Thus, this assumption is satisfied.
However, observations violate the assumption of independence. Children
can be double-counted in the admission data if they are transferred to
multiple different juvenile detention facilities. This means that
juveniles that committed violent crimes may be over-represented because
violent offenders are likely to be in the detention system for longer
thus having a better chance of being moved around. Another assumption of
logistic regression is that the relationship between the probability of
success (in this case the probability of a offender being violent) and
explanatory variables is through the logit function. One way to show
this relation is linear to so prove that interaction terms are not
useful in this model. As previously stated, a Wald test reported a
p-value of
`r round(summary(modelInteraction)$coefficients[[4,4]],digits = 2)`,
meaning that the relationship is truly linear and this assumption is
met.

There are some further limitations to the study, however. First, data is
less widely available for the tail end of juvenile age, so claims about
extremely young or old offenders are limited. Second, there were some
manual entry errors noted in the data for things like years, and it can
be assumed that there were some manual entry errors for variables of
interest. However, because of the mass quantity of the data it is
unlikely that these few errors would have a major impact on the
conclusions in this report. Finally, we note that this data can only be
applied to juveniles in Iowa, as this is where the data comes from. Therefore, the
conclusion from this report is that among children in Iowa housed at
juvenile detention facilities, the proportion that committed violent
crimes is higher among females and decreases as a child grows older.
